---
title: "Is there any inconsistent individual fixel masks?"
author: "Chenying Zhao"
output: html_document
---

This is to test whether there is non-overlapping masks in fixel images. We use the PNC fixel data here (HDF5 file, n=938). This HDF5 file was generated by a recent version of ConFixel (should be "f5a77666").

```{r}
rm(list=ls())
library(devtools)
library(tictoc)

```


```{r}
# install ModelArray with recent version:
ModelArray_commitSHA <- "b99c274c4df2caa77549f58c110e1761955be839"

devtools::install_github(paste0("PennLINC/ModelArray@", ModelArray_commitSHA), 
                         upgrade = "never",  # not to upgrade package dependencies
                         force=FALSE)   # force re-install ModelArray again or not
library(ModelArray)
```

```{r files}
num.subj <- 938
fn.h5 <- paste0("../../data/data_from_josiane/ltn_FDC_n",toString(num.subj), "_confixel.h5")
scalar <- "FDC"
```

Now we create a ModelArray-class object with the .h5 file:
```{r}
modelarray <- ModelArray(fn.h5, scalar_types = scalar)
scalar_matrix <- scalars(modelarray)[[scalar]]
num.elements <- numElementsTotal(modelarray, scalar_name = scalar)
```

First we show that with function `is.finite()`, we can detect if there is any NA/NaN/Inf:
```{r}
is.finite(NA)
is.finite(NaN)
is.finite(Inf)
```

Iterate across all fixels, and check if there is any NA/NaN/Inf:
```{r}
tic.clearlog()
tic("iterating...")

count.not.finite <- 0
for (i_element in 1:num.elements) {
  if (i_element%%1000 == 0) {
    message(paste0("now running element #", toString(i_element)))
  }
  
  values <- scalars(modelarray)[[scalar]][i_element,]
  
  ## check number of subjects with (in)valid values:
  flag_sufficient <- NULL   # whether number of subjects with valid values are sufficient
  num.subj.valid <- length(values[is.finite(values)])
  
  if (num.subj.valid != num.subj) {
    count.not.finite <- count.not.finite + 1
    message(paste0("Element ", toString(i_element), " value is not all finites!"))
  }
}

message(paste0("We detected ", toString(count.not.finite), " elements whose values are not all finites!"))

toc(log=TRUE)  # "iterating..."
```


It says:
We detected 0 elements whose values are not all finites!

iterating...: 11079.438 sec elapsed
