---
title: "Does representation of sex affect the statistical results?"
author: "Chenying Zhao"
output: html_document
---

This is to test how different representation of sex affects the statistical results. For example, using 1/2 vs using M/F.

## GAM, without interaction term
```{r}
rm(list=ls())
library(dplyr)   # %>%
library(mgcv)
```

Create fake data where:

* column `random` is from 1 and 2
* column `randomMF` is male or female, not ordered
* column `ordered_random` is male or female, with female as reference level
  * reference: https://bart-larsen.github.io/GAMM-Tutorial/

```{r}
rm(list=ls())
fn.csv.test <- "../../ModelArray/inst/extdata/n50_cohort.csv"
# this `phenotypes` is a fake data only for test purpose!
phenotypes <- read.csv(fn.csv.test)

# sex in 1 vs 2 (M vs. F):
phenotypes$random <- c(rep(1,25),rep(2,25))
# sex in m vs f:
phenotypes$randomMF <- c(rep("male",25),rep("female",25))
# sex in m vs f, ordered, f as reference group:
phenotypes$ordered_randomMF <- ordered(phenotypes$randomMF, levels = c("female","male"))

head(phenotypes)
```

```{r}
mygam <- mgcv::gam(data=phenotypes, formula=age~factorB + random)
mygam_mf <- mgcv::gam(data=phenotypes, formula = age ~ factorB + randomMF)
mygam_mf_ordered <- mgcv::gam(data=phenotypes, formula = age ~ factorB + ordered_randomMF)

message("GAM with 1 and 2 for sex:")
mygam %>% summary()

message("GAM with m and f for sex:")
mygam_mf %>% summary()

message("GAM with ordered male/female for sex, with female as ref group:")
mygam_mf_ordered %>% summary()
```


Therefore, Using 1/2 or M/F or ordered M/F, GAM models:

* For term sex:
  * p-value: same
  * t value: absolute value is the same, but can be flipped (with DIFFERENT sign), depending on which group is referenced
  * estimate, std.error: DIFFERENT between ordered vs not ordered!
* For terms other than sex and intercept, e.g., factorB here:
  * same
* For model's stat (displayed in `summary()`):
  * same
* For term Intercept:
  * DIFFERENT!

