---
title: "plotting GAM"
author: "Chenying Zhao"
output: html_document
---

This is to test out different methods for plotting GAM

## Bart's vs Graham's
* Bart's GAMM_plotting.R
* Graham Baum's [tutorial](https://github.com/PennBBL/tutorials/wiki/GAM-Syntax-and-Visualization) and [code](https://github.com/PennBBL/tutorials/blob/master/code/gam/GAM_tutorial.R)

```{r set up}
rm(list=ls())

require(ggplot2)
require(mgcv)
library(visreg)   # Graham's code
library(dplyr)  # %>%
```


```{r Graham Baum code for GAM plotting}
# this is copied from: https://github.com/PennBBL/tutorials/blob/master/code/gam/GAM_tutorial.R
# modified so that it behaves as a function
# comments were added by Chenying

#' @param xvar A character, variable name to be plotted on x-axis
Baum_plot <- function(gamModel, xvar, xlabel = NULL, ylabel = NULL) {
  
  # see: https://pbreheny.github.io/visreg/reference/visreg.html
  plotdata <- visreg(gamModel, xvar,
                     data = gamModel$model,     # original Baum's code does not have this; have to add it to avoid error
                     type = "conditional",  # If 'conditional' is selected, the plot returned shows the value of the variable on the x-axis and the change in response on the y-axis, holding all other variables constant (by default, median for numeric variables and most common category for factors). 
                     scale = "linear",  # ?????
                     plot = FALSE)
  
  smooths <- data.frame(Variable = plotdata$meta$x, 
                        x=plotdata$fit[[plotdata$meta$x]], 
                        smooth=plotdata$fit$visregFit, 
                        lower=plotdata$fit$visregLwr, 
                        upper=plotdata$fit$visregUpr)
  predicts <- data.frame(Variable = "dim1", 
                         x=plotdata$res[[xvar]],
                         y=plotdata$res$visregRes)
  
  theplot <- ggplot() +
    geom_point(data = predicts, aes(x, y), colour = "darksalmon", alpha=0.7, size = 1.6 ) +
    geom_line(data = smooths, aes(x = x, y = smooth), colour = "midnightblue",size=2) +
    geom_line(data = smooths, aes(x = x, y=lower), linetype="dashed", colour = "midnightblue", alpha = 0.9, size = 0.9) + 
    geom_line(data = smooths, aes(x = x, y=upper), linetype="dashed",colour = "midnightblue", alpha = 0.9, size = 0.9) +
    theme(legend.position = "none") +
    labs(x = xlabel, y = ylabel) +
    theme(axis.title.x = element_text(size = rel(1.6))) +
    theme(axis.title.y = element_text(size = rel(1.6))) + 
    theme(axis.text = element_text(size = rel(1.4))) + theme(axis.line = element_line(colour = 'black', size = 1.5), axis.ticks.length = unit(.25, "cm")) +
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_blank())
  
  #dev.off()

  # return
  theplot
}

```


My understanding of Bart Larsen's code for plotting GAM: in the case of smooth term only (without interaction term): we will only use the chunk on and after line 281 "No interaction variable, just produce a single line plot".

* scatter plot: use original data from `themodel$model`
* fitted plot + confidence interval: predicts with data:
  * number of data = np = 10,000 (default)
  * x axis (i.e. smooth term age): sample out np data within [min, max] of age
  * other covariates: 
    * for numeric: use median value (for numeric)
    * for factors or ordered: use the levels(df[,thisVar])[[1]], which my understanding is the reference level (i.e. the level #1)
  * y axis: predicted with above data + the input model

For my case, as sex is numeric (1 or 2), it will take median but it will be a integer 2 (number of female is more than male)

```{r Bart code for plotting GAM}
source("../../ModelArray_paper/notebooks/GAMM_plotting.R")   # get `visualize_model()`
```

We load some data for testing - this data is for ModelArray paper figure 5B. How this data was generated: run `analyze_signi_cluster.Rmd` from ModelArray_paper repo with settings:

* ModelArray_commitSHA = "0911c4ffbcc737ea9a615f7a663f57bb0b4e174d" 
* input h5 file: ltn_FDC_n938_wResults_nfixels-0_20220109-183909.h5
* threshold: s_Age.p.value < 1e-15
* formula <- FDC ~ s(Age, k=4, fx=TRUE) + sex + dti64MeanRelRMS
* method.gam.refit <- "REML" 

then save it with:
`save(results, file = "/home/chenying/Desktop/fixel_project/data/data_from_josiane/results/ltn_FDC_n938_wResults_nfixels-0_20220109-183909_forPlottingCluster.RData")`

```{r load data}
fn.rdata <- "/home/chenying/Desktop/fixel_project/data/data_from_josiane/results/ltn_FDC_n938_wResults_nfixels-0_20220109-183909_forPlottingCluster.RData"

load(fn.rdata)  # variable: results + df_avgFixel
# results$f: the figure from Bart's code
# results$onemodel: the fitted model for this averaged FDC from this cluster
# results$pred: the data for drawing the fitted line + se boundaries
# df_avgFixel: the original data

onemodel <- results$onemodel
pred_avgFixel <- results$pred

# a little bit information about the fitted line:
message("the value used for plotting fitted line:")
message(paste0("sex = ", unique(pred_avgFixel$sex) %>% toString()))
message(paste0("motion = ", unique(pred_avgFixel$dti64MeanRelRMS) %>% toString() ))

if ((unique(pred_avgFixel$dti64MeanRelRMS)) == (df_avgFixel$dti64MeanRelRMS %>% median())) {
  message("which is the median value of motion in the original data")
}
```




```{r}
baumplot <- Baum_plot(onemodel, "Age", "Age (years)","FDC")
```

Next steps: to extract 

* the plotted scatter dots
* fitted lines + confidence intervals

to compare...